---
title: "Solutions to the Univariate Models assignment"
output: pdf_document
---
## Univariate Assignment

```{r, echo=FALSE}
# setup the R enviornment for kniting markdown doc properly
library(knitr)
opts_knit$set(root.dir='../', tidy = TRUE)
```

Read in tree data, metadata can be found [here](../data/tree_metadata.txt).

```{r read in pkgs and data}
library(car)
library(MASS)
library(GGally)
source('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/scripts/utility_functions.R')

# read in data
maple <- read.csv('./data/maple.csv', stringsAsFactors = TRUE)
fir <- read.csv('./data/fir.csv', stringsAsFactors = TRUE)
```

1. Carry out an exploratory analysis using the tree dataset. 
Develop and compare models for species cover for a habitat generalist
[_*Acer rubrum*_ (Red maple)](http://www.durhamtownship.com/blog-archives/pix/November1407.jpg) 
and a habitat specialist [_*Abies fraseri*_ (Frasier fir)](https://upload.wikimedia.org/wikipedia/commons/d/d0/Abies_fraseri_Mitchell.jpg). 
Because this dataset includes both continuous and discrete explanatory
variables use the function `Anova` in the packages `car`. After loading `car`
we can call the function like so:

    ```{r, eval=FALSE}
    Anova(my_mod, type=3)
    ```

    This will estimate partial effect sizes, variance explained, and p-values for 
    each explanatory variable included in the model. 

    Compare the p-values you observe using the function `Anova` to those generated
    using `summary`. 

    For each species address the following additional questions:

        * how well does the exploratory model appear to explain cover?
        * which explanatory variables are the most important?
        * do model diagnostics indicate any problems with violations of
          OLS assumptions?
        * are you able to explain variance in one species better than another?


```{r pairs, warning=FALSE}
# prior to model fitting I will visually examine correlations with cover
pairs(maple, lower.panel = panel.cor, upper.panel = panel.smooth)
pairs(fir, lower.panel = panel.cor, upper.panel = panel.smooth)
``` 

The `GGally` package also has a slick ggplot option for pairs plots
```{r ggpairs}
ggpairs(maple)
ggpairs(fir)
```

These plots already tell us the take home message in many ways which is that red
maple cover is not well explained by the measured enviornmental variables with
the exception of elevation. Fraser fir (*Abies fraseri*) may correlate better
with the environmental variables but it is difficult to tell because it has a
cover of zero in so many plots.

Now we will carry out OLS model fitting. As this is an explanatory modeling
we will include all of the variables in the models. We do this for several 
reasons: 

1. this is the best way to fairly compete the variables against one another
2. a partial correlation between the response and an explanatory variable may
be stronger than its raw correlation due to interference from other variables.

```{r ols models}
# build OLS models using "." shorthand which is short for all variables in data
maple_lm <- lm(cover ~ . , data=maple)
fir_lm <- lm(cover ~ . , data=fir)
```

Let's first examine the overall model fit by examining the predicted-observed (PO) plot: 

```{r lm observed vs predicted plots}
par(mfrow=c(1, 2))
plot(predict(maple_lm), maple$cover, xlab='Predicted cover', ylab='Observed cover')
abline(a=0, b=1)
plot(predict(fir_lm), fir$cover, xlab='Predicted cover', ylab='Observed cover')
abline(a=0, b=1)
par(mfrow=c(1, 1))
```

There are some pretty large and systematic deviations from the one-to-one line
which suggests that our model is performing poorly for both species. Additionally
it is obvious that observed cover values are integers while the predicted values
are continuous. Negative cover is predicted for fir but this doesn't
make any sense. We'll return to this idea later when we fit the Poisson GLM
model which only predicts positive integer values.


```{r lm diagnostics}
# before diving into statistics we should check model diagonostics
par(mfrow=c(2,2))
plot(maple_lm)
plot(fir_lm)
par(mfrow=c(1,1))
```

Those diagnostic plots are waving some red flags. In particular, the "Residuals 
vs Fitted" and "Scale-Location" plots indicate that their are some systematic 
deviations of the residuals. You should not see regular geometric patterns in 
residuals so that is a bit troubling. These systematic errors are occurring
because the response `cover` is a discrete variable and the Gaussian error term
in our OLS assumes a continuous response. So there is a mismatch between our
error distribution and the variable we are modeling.


Partial regression plots provide another useful visual representation of the
multiple regression model:
```{r partial regression}
par(mfrow=c(2,3))
termplot(maple_lm, se = TRUE, partial.resid = TRUE)
par(mfrow=c(2,3))
termplot(fir_lm, se = TRUE, partial.resid = TRUE)
```

Above the two sets of plots are displaying the partial residuals, best fit line,
and the 95% confidence interval for each term in the maple and fir multiple
regression models respectively. You can think of these relationships as the
correlation between the variable and the response after the other variables have
been statistically controlled for (i.e., the partial effect).

From these plots we can see that elevation has the strongest effect on both
trees where it is negative for maple and positive for fir. The tci variable
appears to be the next most important variable where maple has a negative
relationsihp and fir has a positive relationship. The other variables appear to
be a wash for the most part.

The function `termplot` can be used to examine partial effects for `glm` models
as well. The `car` package also provides a nice function for examining partial
residuals called `crPlots` (see
[here](https://www.statology.org/partial-residual-plot-in-r/) for examples).

Now let's examine the output from the functions `summary` and `Anova` to examine
    1. partial variable importance (t-value effect size and significance)
    2. overall model performance (adjusted r-squared and significance)

```{r maple lm tables}
summary(maple_lm)
Anova(maple_lm, type=3)
```

Overall the model for maple explained `r round(summary(maple_lm)$adj.r.squared, 2) * 100`% of the variance in cover based upon the computed adjusted $R^2$. 
This actually would not necessarily be considered a terrible $R^2$ value if it were not for the systematic deviations we already observed in the fitted and residual
values of the model which indicate a fundamentally flawed model. 

The strongest correlations for this species are negative with respect to `elev`, `tci` and `beers` as evidenced by their negative t-values which were significantly different than zero. 
This indicates that this species prefers lower elevations that are dryer and warmer.

```{r fir lm tables}
summary(fir_lm)
Anova(fir_lm, type=3)
```

For fir the results are a bit different. This species is slightly better
explained by the explanatory variables (adjusted $R^2$ = 
`r round(summary(fir_lm)$adj.r.squared, 2)`). 

The ANOVA table indicates that this species is primarily driven by `elev` with a
secondary role for `disturbance`. Note the F-statistics for those variables are
much larger. The `summary` output which provides the coefficient estimates
suggests that this species prefers high elevations (positive regression
coefficient) and VIRGIN forests (the most positive \(\beta\) of the disturbance categories)
- remember these \(\beta\)'s are relative to the intercept which in this case is the disturbance category `r levels(fir$disturb)[1]`).

We were able to model the habitat specialist, fir, about as well as
the more generalist species maple, but we have some serious concerns and
doubts about our inferences because the model diagnostics and common sense
suggest that a Gaussian error for this response variables is inappropriate.


2. You may have noticed that the variable cover is defined as positive integers
between 1 and 10. and is therefore better treated as a discrete rather than
continuous variable. Re-examine your solutions to the question above but from
the perspective of a General Linear Model (GLM) with a Poisson error term 
(rather than a Gaussian one as in OLS). The Poisson distribution generates
integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

    ```{r, eval=FALSE}
    maple_poi <- glm(cover ~ tci + elev + ... , data= my_data, 
                   family='poisson')
    ```

    For assessing the degree of variation explained you can use a 
    pseudo-R-squared statistic (note this is just one of many possible)

    ```{r}
    pseudo_r2 <- function(glm_mod) {
        1 -  glm_mod$deviance / glm_mod$null.deviance
    }
    ```

    Compare your qualative assessment of which variables were most important in each model. 
    Does it appear that changing the error distribution changed the
    results much? In what ways? 

```{r poisson regression}
maple_poi <- glm(cover ~ . , data = maple, family = 'poisson')
fir_poi <- glm(cover ~ . , data = fir, family = 'poisson')
```

Let's examine the predicted-observed plots first to see if they systematic error
is better handled.

```{r poisson PO plot}
par(mfrow=c(1, 2))
plot(predict(maple_poi, type='response'), maple$cover,
     xlab='Predicted cover', ylab='Observed cover')
abline(a=0, b=1)
plot(predict(fir_poi, type='response'), fir$cover,
     xlab='Predicted cover', ylab='Observed cover')
abline(a=0, b=1)
par(mfrow=c(1, 1))
```

These PO plots already look much more encouraging.
There is still quite a bit of error but it is less systematic and 
their are no predicted negative cover values. 
Let's look at the model diagnostics:

```{r poisson model diagnostics}
par(mfrow=c(2,2))
plot(maple_poi)
plot(fir_poi)
par(mfrow=c(1,1))
```

These model diagnostics look better but still not great. We can still observe
some systematic patterns of error in the residuals plots. Also the Fraser fir
residuals are still decidedly non-Normal.
This seems to be due in part to the large number of 
samples with zero cover of *Abies fraserii*. 

One important thing to note is that the predicted and residual values in the
above plots are on the transformed log scale. It can be more intuitive and
informative to plot them on the response scale as follows. 

```{r}
par(mfrow=c(1,2))
plot(predict(maple_poi, type='response'),
     residuals(maple_poi, type='response'), 
     xlab='Fitted values', ylab='Residuals',
     main='maple')
plot(predict(fir_poi, type='response'),
     residuals(fir_poi, type='response'),
     xlab='Fitted values', ylab='Residuals',
     main='fir')
par(mfrow=c(1,1))
```

Before we examine differences in the model outputs, let's examine if using the
Poisson error term improved model fit.

```{r}
summary(maple_lm)$r.squared
pseudo_r2(maple_poi)
summary(fir_lm)$r.squared
pseudo_r2(fir_poi)
```

For red maple the fit is worse but for Fraser fir we observe a big improvement
in fit. This is in large part due to the fact that the Poisson distribution is
truncated at zero and given that for Fraser fir there are a lot of sites with
zero cover where the Gaussian model was doing a very poor job.

The PO plots For both Poisson models suggest that the models tend to under
predict the occurrence of the zero category. This is most pronounced for Fraser
fir. So even though we've explicitly recognized the positive discrete nature of
our response variable it is still inflated with zeros relative to a Poisson
distribution.

Given that the Poisson regression actually performed worse for the Red maple
than the OLS regression. I'll not interpret that model further.

```{r}
summary(fir_poi)
Anova(fir_poi, type=3)
```

The take home messages we arrived at with the OLS modeling have not changed
greatly for Fraser fir. More variables are statistically significant but the
real story remains about the strong effect of elevation and disturbance.

At this point we may still not be very satisfied with our modeling though given
the excess of zeros we observed above the Fraser firs. One option would be to
try to use a negative binomial error term rather than Poisson which provides for
greater aggregation of zeros due to the inclusion of an additional clumping
parameter.

```{r fir negative binomial}
library(MASS)

fir_nb <- glm.nb(cover ~ . , data=fir,
                  control=glm.control(maxit=100))
AIC(fir_poi)
AIC(fir_nb)
```

That resulted in a modest decrease in the AIC. One last model to examine is a
zero inflated Poisson model in which combines two models: 

* a logistic regression for whether or not cover is zero or not, and 
* a Poisson regression for variation in the size of cover. 

```{r fir zip}
library(pscl)

# fit a model in which all the variables are included
# in the portion of the model with positive values, and 
# only include elevation in the model for the zeros. 

fir_zip <- zeroinfl(cover ~ . | elev, data=fir)

AIC(fir_poi)
AIC(fir_nb)
AIC(fir_zip)

# PO plot
plot(predict(fir_zip), fir$cover)
abline(a=0, b=1)

# diagonstic plot
plot(predict(fir_zip), residuals(fir_zip))
```

This model resulted in a substantial increase in the model adequacy as judged by
the much lower AIC. The model diagnostic plot looks more reasonable and when we
examine the predicted to observed plot we see that we are doing a better job
predicting all the zeros in the dataset.

```{r}
Anova(fir_zip, type=3)
```

The above output from `Anova` isn't so helpful. We see that `disturb` is marginally important but in this case `summary` provides more useful information. 

```{r}
summary(fir_zip)
```

Above we see the coefficient estimates for the Poisson portion of the model
first and then below the coefficients for the logistic portion of the model.
Elevation had a very important role on the performance of the logistic
regression which isn't a big surprise based on the very first graph we made
using `pairs`. Elevation likely also was relevant for the Poisson portion based
on what we've learned with the other models but it is difficult to be sure
because the estimate of standard error, the z-value, and the p-value are missing
for this variable. This is because of the warning that was generated with
summarizing the model output. I'm not exactly sure what happened but when you
look at the variance-covariance matrix the diagonal element for elevation is
very small and the sqrt of a very small number is NaN. This must be needed for
downstream estimates of standard error and the like. To round out our impression
of our ability to model the influence of elevation on *Abies frasier* cover we
can plot our predictions against the data. To do so we'll make it a bit easier
by only considering a zero-inflated Poisson model that only includes elevation.

```{r}
fir_elev_gau <- glm(cover ~ elev, data=fir, family='gaussian')
fir_elev_poi <- glm(cover ~ elev, data=fir, family='poisson')
fir_elev_zip <- zeroinfl(cover ~ elev | elev, data=fir)

fir_newdata <- data.frame(elev = seq(min(fir$elev), max(fir$elev),
                                 length.out=100))
plot(cover ~ elev, data=fir, xlab='Elevation', 
     ylab='cover', main='fir',
     ylim=c(-1, 10))
lines(lowess(fir$elev, fir$cover), col='red', lwd=2)
lines(fir_newdata$elev,
      predict(fir_elev_gau, newdata=fir_newdata,
              type='response'), col='dodgerblue', lwd=2)
lines(fir_newdata$elev,
      predict(fir_elev_poi, newdata=fir_newdata,
              type='response'),  col='purple3', lwd=2)
lines(fir_newdata$elev, 
      predict(fir_elev_zip, newdata=fir_newdata),
      col='blue', lwd=2)
legend('topleft', c('lowess', 'Gaussian', 'Poisson', 'zero-inflated Poisson'), 
       col=c('red', 'dodgerblue', 'purple3', 'blue', 'green'),
       lwd=2, bty='n')
```

Let's examine if the zero-inflated Poisson improves our maple model as well.

```{r}
maple_zip <- zeroinfl(cover ~ . | elev, data=maple)

AIC(maple_poi)
AIC(maple_zip)

summary(maple_zip)

plot(predict(maple_zip), maple$cover)
abline(a=0, b=1)
plot(predict(maple_zip), residuals(maple_lm))



maple_elev_gau <- glm(cover ~ elev, data=maple, family='gaussian')
maple_elev_poi <- glm(cover ~ elev, data=maple, family='poisson')
maple_elev_zip <- zeroinfl(cover ~ elev | elev, data=maple)

maple_newdata <- data.frame(elev = seq(min(maple$elev), max(maple$elev),
                                 length.out=100))

plot(cover ~ elev, data=maple, xlab='Elevation', 
     ylab='cover', main='maple',
     ylim=c(-1, 10), xlim=c(250, 2250))
lines(lowess(maple$elev, maple$cover), col='red', lwd=2)
lines(maple_newdata$elev,
      predict(maple_elev_gau, newdata=maple_newdata,
              type='response'), col='dodgerblue', lwd=2)
lines(maple_newdata$elev,
      predict(maple_elev_poi, newdata=maple_newdata,
              type='response'),  col='purple3', lwd=2)
lines(maple_newdata$elev, 
      predict(maple_elev_zip, newdata=maple_newdata),
      col='blue', lwd=2)
legend('topright', c('lowess', 'Gaussian', 'Poisson', 'ZIP'), 
       col=c('red', 'dodgerblue', 'purple3', 'blue', 'green'),
       lwd=2, bty='n')
```

The payoff is not as great for adopting a ZIP model for maple, but the
AIC is still substantially lower and the diagnostic residual plot is much
better behaved.

This last plot for maple suggests a uni-modal response of 
this species with elevation which could either be best captured
using a weighted averaging approach of including a quadratic elevation term into the model (i.e., `elev^2`). 

3. Provide a plain English summary (i.e., no statistics) of what you have
found and what conclusions we can take away from your analysis?

The take home messages from this analysis are that both species are responding 
to the environment although fir which is more of a habitat 
specialist shows stronger correlations with the available enviornmental 
variables. Elevation was the most important variable in all the models we 
examined and ecologically this is not a big surprise either given its combined
influence on moisture and temperature. From a modeling perspective we also
gained some insight when working with discrete data. If there are a lot of zeros
we observed that developing a separate model for the zeros was extremely
beneficial for constructing more accurate and reasonable model predictions.
Interestingly though the take home messages of which variables were important
did not change greatly from our initial inference from the OLS models. This
indicates OLS is fairly robust to substantial violations of its assumptions.
Most importantly it is key to recognize that  lot of the insight we gained from
the model analysis was visually pretty obvious from our initial observation of
the data patterns. This is a good reminder to always plot the data!

4. (optional) Examine the behavior of the function `step()` using the 
exploratory models developed above. This is a very simple and not very robust
machine learning stepwise algorithm that uses AIC to select a best model. By
default it does a backward selection routine.

```{r optional stepwise models}
maple_step <- step(maple_lm)
# streamdist is dropped
summary(maple_step)
fir_step <- step(fir_lm)
summary(fir_step)
# beers is dropped
```
In the code above, I have provided the methods for using a backward stepwise
algorithim to reduce the model complexity when the input model is an OLS model
fit with `lm`. You can see that the function `step` outputs information about
increased model fit as the algorithim removes variables (i.e., works backward)
For both the `maple_lm` and `fir_lm` objects the algo only removes a single
variable before setting on what is consider the most parsimonious model (i.e.,
the lowest AIC). In this particular case we really do not learn anything new as
this effectively results in just dropping the non-significant variables form the
models which we already had a good clue were unimportant. The stepwise approach
can also be used with the models fit with other model specifications such as
`glm`, `glm.nb` or `zeroinfl`. The `step` function can also be setup to be 
a foward selection algorithim using its `scope` and `direction` arguments. As noted in the univariate lesson, automated variable selection approaches that 
compare many different models should be treated as exploratory and not for
strict hypothesis testing. 

